Denotational Semantics of General Payment Primitives, and Its Payment System

### 2022-09-29 packages/spec-haskell/yellowpaper/Paper.tex
Denotational Semantics WIP!

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Document Setup
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% This document is processed in an isolated clean folder at the same level.
\makeatletter
\def\input@path{{../tex/}}
\makeatother
%%% Include the preamble at choice during processing.
\input{preamble.tex}
%%% Bibliography system setup using biblatex.
\usepackage[
    backend=biber,
    style=alphabetic,
]{biblatex}
\addbibresource{../tex/Biblio.bib}
%%% Macros used.
\long\def\ignore#1{} % for ignoring some code
\newcommand{\eg}{\textit{e}.\textit{g}.}
\newcommand{\ie}{\textit{i}.\textit{e}.}
\newcommand{\etal}{\textit{et al}.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Document Meta Data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\author{\\
    Miao, ZhiCheng\\
    Co-Founder, Superfluid Finance\\
    miao@superfluid.finance
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Document Body
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\maketitle

@@

This discrete nature of this money distribution model is schematically depicted in \ref{fig:discrete-md}:

@@

\begin{itemize}
\item If time (t) is included the context, then monetary value of each monetary unit can be continuous over time. Time
    is part of the physical reality, hence not changeable by the actors in the payment system.
\item Any subset of monetary units can also have their value functions depending on the same information in
    context. This could enable payment primitives that involve many parties. This set of information in context is
    referred to as \textit{ctx :: SharedContext}, they can be changed over time by the actors in the payment system.
\end{itemize}

@@

@@

Here is the specification of denotational semantics that support both discrete and continous payments.

Here is the list of acronyms used:

\begin{itemize}
    \item \textbf{MD} is \textit{Money Distribution}.
    \item \textbf{MU} is \textit{Monetary Unit}.
    \item \textbf{RTB} is \textit{Real Time Balance}.
\end{itemize}

\subsection{Generalized Payment Model}

\begin{equation}\label{sem_transfer}
    \begin{split}
        [\![MD\ mu\ t\ rtb]\!] = mu \rightarrow t \rightarrow rtb
    \end{split}
\end{equation}

\subsection{Monoidial Payment Model}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{equation}\label{sem_mzero}
    [\![\emptyset]\!] = \lambda\ mu\ t\ \rightarrow 0
\end{equation}

\begin{equation}\label{sem_mappend}
    [\![mda \oplus\ mdb]\!] = \lambda\ mu\ t\ \rightarrow
    [\![mda]\!]\ mu\ t\ +\ [\![mdb]\!]\ mu\ t
\end{equation}

@@

@@

@@

\section{Restatement in Agda and Correctness Proof}

>>>>

%% Include a preamble at choice during processing.
\input{preamble.tex}
\begin{document}

@@

\author{\\
    Miao, ZhiCheng\\
    Co-founder, Superfluid Finance\\
    miao@superfluid.finance
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title Page
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\maketitle

@@

This discrete nature of this money distribution model is schematically depicted in~\ref{fig:discrete-md}:

@@

\begin{itemize}
\item If time (t) is included the context, then monetary value of each monetary unit can vary continuously over
    time. Time is part of the physical reality, hence not changeable by the actors in the payment system.
\item Any subset of monetary units can also have their value functions depending on the same information in
    context. This could enable payment primitives that involve many parties. This set of information in context is
    referred to as \textit{ctx :: SharedContext}, they can be changed over time by the actors in the payment system.
\end{itemize}

@@

The set of payment primitives supported in a payment system is general, if monetary value of each monetary unit can vary
continuously over time and monetary values can logically be shifted between two or more monetary units. The extent of
the generality of each payment system may vary. We introduce a set of payment primitives that is general and serves as a
starting point for the readers to explore the space of modern payment system. The specifications will be formally
defined using denotational semantics, and its restatement in Haskell is also applied as a constructive artifact friendly
to computer environment\footnote{One may argue for a restatement in \textit{Agda} instead, for it has a richer
dependently-type system for needed constructive proofs. This matter will be dealt in the ``future investigations'' chapter
of this paper.}.

@@

Here is the convention for symbols used in the formulas:

\begin{itemize}
    \item \textbf{M} is \textit{money distribution}.
    \item \textbf{U} is for \textit{the set of all money units in M}.
    \item $\boldsymbol{\nu}$ is for \textit{monetary value}.
    \item \textbf{t} is for \textit{time}.
    \item \textbf{mu, ma, mb, etc.} are for \textit{monetary units}.
\end{itemize}

Recall the definition of payment system previously, money distribution sits in the core of the system, and that is the
syntactic category $\mathcal{M}$ we will be dealing with:

\begin{itemize}
\item The meaning of $[\![\mathcal{M}]\!]$ is money distribution.
\item Semantic function $[\![.]\!] :: \mathcal{M} \rightarrow [\![\mathcal{M}]\!] $ include payment primitives.
\end{itemize}

\paragraph{Model}

This is the model for money distribution.

\begin{equation}\label{md_model}
    \begin{split}
        [\![M\ mu\ t\ \nu]\!] &= mu \rightarrow t \rightarrow \nu \\
        [\![.]\!] &= M\ mu\ t\ \nu \rightarrow (mu \rightarrow t \rightarrow \nu) \\
    \end{split}
\end{equation}

\paragraph{Monoidal}

With this model, $\mathcal{M}$ is also monoidal, hence compositional.

\begin{equation}
    \begin{split}
        [\![\emptyset]\!] &= \lambda mu \times t\ \rightarrow 0 \\
        [\![ma \oplus mb]\!] &= \lambda mu \times t\ \rightarrow [\![ma]\!]\ mu\ t\ +\ [\![mb]\!]\ mu\ t \\
    \end{split}
\end{equation}

Knowing that function application category $a \rightarrow b$ is also monoidal:

\begin{equation}
    \begin{split}
        \emptyset &= \lambda a \rightarrow \emptyset \\
        f \oplus g &= \lambda a \rightarrow f\ a \oplus g\ a \\
    \end{split}
\end{equation}

With some substitutions, we get the desired \textit{monoid homomorphism} property for $\mathcal{M}$.

\begin{equation}
    \begin{split}
        [\![\emptyset]\!] &= \emptyset \\
        [\![ma \oplus mb]\!] &= [\![ma]\!] \oplus [\![mb]\!] \\
    \end{split}
\end{equation}

@@

A payment primitive \textit{prim} is a semantic function that produces a monoidal money distribution, semantically
representing an ``update'' to a money distribution:

\begin{equation}
    [\![prim\ args]\!] = [\![M\ mu\ t\ \nu]\!]
\end{equation}

\paragraph{Law of Conservation of Value}

First of all, money distribution must obey the \textit{law of conservation of value}:

\begin{equation}
    \forall t \in \mathbb{R} {\displaystyle \sum_{mu \in U} [\![M\ mu\ t\ \nu]\!]\ mu\ t = 0}
\end{equation}

That is to say, for any given time, the sum of monetary value of all monetary units shall always equal to zero. Zero is
used to keep the semantical meaning of payment system simple and elegant. In its applications, a payment system may use
some special monetary unit accounting negative monetary values to accommodate concepts such as mining, minting, money
printing, etc.

Further more, in a payment system, we restrict that only payment primitives can provide ``updates'' to its money
distribution. So a money distribution is a result of a sequence of payment primitive updates:

\begin{equation}
    \begin{split}
        &\textbf{Axioms A:\ }
        \textit{law of conservation of value for payment primitives} \\
        &\forall t \in \mathbb{R} ({\displaystyle \sum_{mu \in U} [\![prim\ args]\!]\ mu\ t = 0}) \\
        \\
        &\textbf{Axioms B:\ }
        \textit{money distribution consists of updates from payment primitives} \\
        &[\![M\ mu\ t\ \nu]\!] =
          [\![prim_1\ args_1]\!] \oplus [\![prim_2\ args_2]\!] \oplus [\![prim_3\ args_3]\!] \oplus \dotsb
    \end{split}
\end{equation}

This allows us impose the law of conservation of value on payment primitives, then prove the law for money
distribution inductively using monoid homomorphism:

\begin{equation}
    \begin{split}
        &\textbf{In addition to the axioms, given:} \\
        \\
        &\textit{(base case with empty money distribution is trivial)} \\
        &\forall t \in \mathbb{R} {\displaystyle \sum_{mu \in U} [\![\emptyset]\!]\ mu\ t = 0} \\
        \\
        &\textit{(monoid homomorphism)} \\
        &{\displaystyle \sum_{mu \in U} [\![prim_1\ args_1\oplus prim_2\ args_2]\!]\ mu\ t} = \\
        &\qquad {\displaystyle \sum_{mu \in U} [\![prim_2\ args_2]\!]\ mu\ t} \oplus
          {\displaystyle \sum_{mu \in U} [\![prim_2\ args_2]\!]\ mu\ t} \\
        \\
        &\textbf{We have:} \\
        &\forall t \in \mathbb{R}. \\
        &{\displaystyle \sum_{mu \in U} [\![M\ mu\ t\ \nu]\!]\ mu\ t} \\
        &\textit{(applying Axioms B)} \\
        = &{\displaystyle \sum_{mu \in U}
            ([\![prim_1\ args_1]\!] \oplus
            [\![prim_2\ args_2]\!] \oplus
            [\![prim_3\ args_3]\!]\oplus \dotsb)
        }\ mu\ t\\
        &\textit{(applying monoid homomorphism)} \\
        = &{\displaystyle \sum_{mu \in U} [\![prim_1\ args_1]\!]\ mu\ t} \oplus
        {\displaystyle \sum_{mu \in U} [\![prim_2\ args_2]\!]\ mu\ t} \oplus
        {\displaystyle \sum_{mu \in U} [\![prim_3\ args_3]\!]\ mu\ t} \oplus \dotsb
        )\\
        &\textit{(applying Axioms A)} \\
        =&\ 0
        \\
        \blacksquare
    \end{split}
\end{equation}

\subsection{One to One Payment Primitives}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

???

@@

\subsection{Index Abstraction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Proportional Distribution Primitives}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\section{Restatement in Agda, Correctness and Equivalence Proofs}
